### 2025-05-17 manuscripts/10-JERP-v3/manuscript_main.tex
some edits

A large body of literature provides evidence of substantial earnings gaps across race and ethnicity in the US \autocite{bayer2018divergent, charles2008prejudice}. Hispanic individuals constitute a large and growing portion of the US population, and as this number increases, it is increasingly crucial to determine whether ethnic discrimination affects their employment prospects, wages, and career advancement. Understanding these labor market outcomes is critical, as they directly affect broader societal issues like assimilation and economic mobility, \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017}. These factors serve as key indicators of how successfully Hispanic immigrants can navigate society and ascend the socioeconomic ladder.

In this paper, I answer the following questions. Does having a Hispanic last name affect educational outcomes? Does having a Hispanic last name affect labor market outcomes? I aim to show that comparing Hispanic White individuals to to non-Hispanic White individuals might create an artificially higher earnings gap since the two groups differ in many observable characteristics.\footnote{I define Hispanic individuals as those who self-report their ethnicity as Hispanic on surveys or other data collection instruments.} \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} My analysis focuses on US-born children with foreign-born parent (i.e., inter-ethnic).\footnote{This study's focus on US-born children with foreign-born parents means the findings may not generalize to Hispanic children with US-born parents, who likely face different socioeconomic and cultural circumstances. The analysis does not account for heterogeneity in immigrant characteristics across different Spanish-speaking countries of origin, such as variations in educational attainment, socioeconomic status, and gender-specific migration patterns, which could influence both parental selection into migration and subsequent child outcomes.} Others have attempted to compare how native-born White Hispanics fare compared to non-Hispanic Whites and foreign-born Hispanics. \cites{antman2020ethnic} have compared the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health.\footnote{For additional studies examining ethnic identity and outcomes among Hispanic populations, see \textcite{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, which collectively document the relationships between ethnic identification, assimilation patterns, and various socioeconomic outcomes.} \textcite{davilaChangesRelativeEarnings2008} found gaps in labor market outcomes between Hispanics and Whites, which they attribute largely to differences in education, experience, immigration status, and regional differences. This paper builds on those studies by emphasizing how cultural assimilation and generational status further shape educational and labor market disparities for Hispanics.

Understanding discrimination against Hispanics in labor markets has far-reaching implications that extend beyond individual earnings. Labor market discrimination can create persistent intergenerational disadvantages by limiting economic mobility, reducing access to quality healthcare and education, and constraining residential choices \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,bowles2002inheritance, djajic2003assimilation}. Discrimination may foster occupational sorting and segregation, as shown by recent task-based models of racial wage gaps that identify race-specific barriers in “Contact” tasks and explain the stagnation of these gaps post-1980 \autocite{hurst2024task}. These barriers may trap Hispanic families in cycles of lower socioeconomic status, as reduced earnings and employment opportunities limit their ability to invest in their children's human capital or build wealth through homeownership and savings.\footnote{See \textcite{bowles2002inheritance} for an analysis of the intergenerational transmission of economic status.}

@@

This study builds upon and improves upon the traditional Oaxaca-Blinder-Kitagawa decomposition approach often used in labor market discrimination studies \autocite{kitagawa1955components, oaxaca1973male,blinder1973wage}. While these decomposition methods have provided valuable insights into earnings gaps between groups, including Hispanics and non-Hispanic Whites \autocite{davilaChangesRelativeEarnings2008}, they are limited in their ability to account for unobserved differences between the groups. My approach comparing children of inter-ethnic marriages, provides a more closely matched comparison group, allowing me to better isolate the effect of likely having a Hispanic last name on labor market outcomes.

The methodological improvement in this paper is particularly timely and relevant given the rapidly changing demographics of the United States. The US population is growing increasingly diverse, with significant implications for labor market dynamics and potential discrimination. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019.\footnote{The proportion of non-Whites and Hispanics is based on the author's calculations using the Current Population Survey (CPS).} Native-born White Hispanic men earn 21\% less than White men, although there is evidence that a substantial portion of this gap is due to differences in education \autocite{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. Discrepancies may also be due to discrimination against Hispanics, which can lead to reduced job opportunities, lower wages, and poor assimilation \autocite{chettyWhereLandOpportunity2014,bowles2002inheritance, djajic2003assimilation}.\footnote{This study focuses on US-born children of inter-ethnic unions, many of whom have at least one foreign-born parent. Assimilation concerns remain relevant because foreign-born parents may pass on cultural norms and language preferences, which can still shape these children's educational and labor market outcomes.}

Using this novel approach, I find significant effects of Hispanic surnames on both educational and labor market outcomes, and that individuals that likely have a Hispanic last names face substantial disadvantages in educational attainment and earnings compared to their counterparts with likely non-Hispanic names. Specifically, individuals with presumably Hispanic last names complete 0.2 fewer years of education than those with non-Hispanic names, even when controlling for family background. Individuals with presumably Hispanic last names are also 1 percentage point more likely to be unemployed and earn 5 percentage points less than those with likely non-Hispanic names. These findings suggest that ethnic discrimination continues to play a role in shaping economic opportunities for Hispanics in the United States, highlighting the need for targeted policies to address these persistent disparities.

@@

The main identifying assumption of my empirical strategy depends on the assumption that people born to HW parents are similar to their WH peer in all observable---and unobservable---aspects and characteristics relevant to the labor market, so surname is the only difference between these groups. Children from US born-Hispanic homes might benefit from more favorable family conditions than those from Hispanic-US born homes, considering Whites statistically have higher socioeconomic status than Hispanics.  These factors introduce doubts regarding whether children from Hispanic-US born families receive comparable familial support and influences, genetically or environmentally, as those from US born-Hispanic families.

@@

I use three datasets:  the Integrated Public Use Microdata Series (IPUMS) Current Population Survey (CPS), CPS' Annual Social and Economic (ASEC) supplement, and CPS' outgoing rotation \autocite{cps2019}, and the 1960 to 2000 US censuses \autocite{acs2019}.

I use the 1994–2019 CPS data set to study the effect of Hispanic surnames on labor market outcomes, because that span contains data on parents’ place of birth, ethnicity, and race. The CPS does not provide data on parents’ characteristics, essential to determine the family background, but the 1960–2000 US Census data includes parents’ place of birth, race, and ethnicity. I employ this information to construct "synthetic parents" using a method developed by \textcite{rubinstein2014pride} linking husbands and wives  in the census data. Assuming that parents have children 25–40 years old, I link these “parents” using parents’ places of birth, birth year of "children," and the parents’ places of birth.

@@

I use the CPS for my primary analysis of the effect of a Hispanic surname on earnings. I restrict my sample to US citizens  aged 25–40 born between 1960 and 2000 in the US. Taking advantage of data on parents’ place of birth, I divide the sample into four groups depending on their parents’ ethnicity. Mothers or fathers are Hispanic if they were born in a Spanish-speaking country or Puerto Rico, and US born White if they were born in the United States.\footnote{The list of Spanish-speaking countries includes: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} Therefore, an observation can be the product of four types of parents:
\begin{enumerate}
\item US born father and US born mother (hereafter WW) 
\item US born father and Hispanic mother (hereafter WH)---the comparison group and the group that is likely to have a non-Hispanic-sounding last name
\item Hispanic father and US born mother (hereafter HW)---the focus of this study and the group most likely to have a Hispanic-sounding last name
\item Hispanic father and Hispanic mother (hereafter HH).
\end{enumerate}

@@

@@

Using the 1960 to 2000 censuses, I constructed a data set of synthetic parents. The sample includes married White men and women. Even though the census asks a person whether they are Hispanic or not, I took advantage of the questions on place of birth to create a proxy for ethnicity. I consider a Hispanic persons that self-report White as their racial group and are born in a Spanish-speaking country. Consequently, US born in the sample are those that self-report White as race and are native-born. Using the information provided in the census, I can link husbands and wives with each other. I assume that parents have children between the ages of 25 and 40, so my sample consists of married White men and women with children born in the 1920 to 1975 cohorts\footnote{The construction of ``synthetic parents'' follows the method used by \textcite{rubinstein2014pride}.}.

For example, consider someone who was 35 years old in 1999, meaning they were born in 1964. If this person's mother was born in Mexico and their father was born in the United States, their ``synthetic parents'' would be identified using the 1970 Census, when this person was 6 years old. These ``synthetic parents'' are created by matching the characteristics of Mexican-American couples who had a 6-year-old child in 1970. The ``synthetic mother'' would have the average education level, income, and other socioeconomic characteristics of Mexican-born women who were married to US-born men and had children around 1964, when they were between 20 and 35 years old (meaning they were born between 1929 and 1944). The ``synthetic father'' would have the average characteristics of US-born men who were married to Mexican-born women and had children in that same year. This method preserves the demographic composition of parents while assigning them historically accurate socioeconomic characteristics based on their origins and marriage patterns.

@@

Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variable for the type of parents of person $i$. $X_{ist}$ is a vector of controls that includes age and numbers of hours worked, $\gamma_{st}$ are state-year fixed effects (FE), and $\phi_{ist}$ represents the error term.\footnote{I also include parental characteristics of `synthetic' parents as controls in alternative specifications. These controls include income and education of `synthetic' parents.} The equation for this strategy is written as follows and the sample is restricted to I restrict the sample to WH and HW groups:
\begin{equation} \label{eq:1a}
Y_{ist} = \beta_{1} HW_{ist} + X_{ist} \pi + \gamma_{st} + \phi_{ist}
\end{equation}

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gaps in outcomes between children of inter-ethnic marriages who likely have a Hispanic-sounding last name versus a White-sounding last name. If $\beta_{1} > 0$, then people with a Hispanic last name have better outcomes than people with a White last name. If $\beta_{1} < 0$, then people with a Hispanic last name have worse outcomes than people with a White last name.

By comparing interethnic children, I can control for many unobservable family characteristics that may affect labor market outcomes, since these families are likely to be more similar to each other than to families with two Hispanic or two non-Hispanic White parents. This approach allows me to isolate the effect of surname, providing a more accurate estimate of its impact. The difference in means between Hispanic and non-Hispanic Whites could result from discrimination, but it could also be due to differences in innate abilities, skills, and parental investments . WH children will have a non-Hispanic last name, while HW children will have a Hispanic last name. This method, developed by \textcite{rubinstein2014pride}. This is an improves traditional Oaxaca-Blinder-Kitagawa decomposition methods, which are limited in their ability to account for unobserved differences between groups, by providing a more suitable comparison group to measure discrimination \autocite{oaxaca1973male,blinder1973wage,kitagawa1955components}.

@@

Two primary reasons underscore this assumption. First, there is a significant selection in the marriage market. Since belonging to the Hispanic out-group brings negative societal effects, it is worth exploring factors common between White women that have children with Hispanic men. Second, fathers and mothers influence human capital accumulation differently \autocite{kimball2009risk,magruder2010intergenerational}. If marriage were random, then WH households might be a better environment for children. Using the CPS and the US Census, I will evaluate these empirical concerns.

A large body of empirical and theoretical work indicates that marriages exhibit strong selection, or "assortative matching." \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996, beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}. Also, \textcite{duncanIntermarriageIntergenerationalTransmission2011} show a similar pattern of assortative matching among Mexicans in the United States. Interethnic WH and HW marriages generally consist of partners sharing similar characteristics, suggesting both WH and HW parents are similar in aspects relevant to labor market outcomes.

Another threat to identifying a true effect could arise from measurement error in using parents’ place of birth from the CPS data as a proxy for a parent’s ethnicity and last name: The CPS notes place of birth but not ethnic or racial identity. It is possible that a native-born father could be a second-generation or later immigrant from a Spanish-speaking country, but this is unlikely. Most Hispanics from 1960 to 2000  were first-generation immigrants, and the number of second-generation or later was very small (Table \ref{tab:mat3}).\footnote{Based on the author's calculations using US Census data.} Only 3\% of native-born Americans identified as Hispanic, making it unlikely that an interethnic child with a native-born father is also a second generation Hispanic immigrant.

@@

It is important to acknowledge potential threats to identification in this approach. While HW and WH couples exhibit similar observable socioeconomic characteristics, they may differ systematically in unobservable dimensions that influence labor market outcomes for their children. These unobservable differences could include parenting styles, educational preferences, time investments in children, gender norms, or cultural attitudes that are not captured in conventional socioeconomic metrics. However, this approach offers distinct advantages over traditional Oaxaca-Blinder-Kitagawa decomposition methods. First, by focusing specifically on children from interethnic marriages with similar parental characteristics but different surname ethnicities, I use a more targeted comparison group that better isolates the effect of perceived ethnicity from family background. Traditional decomposition approaches would struggle to separate discrimination effects from the broader socioeconomic gaps between Hispanic and White families overall. Second, the smaller educational and income disparities between HW and WH families compared to HH and WW families suggest that selection bias is less severe in this comparison, allowing for a clearer attribution of outcome differences to discrimination rather than unobserved heterogeneity. Nevertheless, I acknowledge that residual selection bias remains a concern, particularly regarding cultural and attitudinal factors that may systematically differ between HW and WH households despite similar educational and economic profiles.

Using the “synthetic” parents I constructed using the Census data, I examine the family background of the different types of children. WW couples have the highest level of education among the four groups: 12.58 years  for husbands and 12.36 for wives, so a WW household has 24.95 years of schooling. Men in HH marriages have 8.91 years of education, while women have 8.68. As a household, HH couples have 17.69 years of education. As predicted, interethnic couples marry people similar to them. WH husbands have 11.82 years of education, while wives have an average of 10.71, a household total of 22.68 years. HW husbands have 10.33 years, while wives have an average of 11.01, giving HW households a total of 21.50 years. Both  Hispanic men and women in interethnic couples marry white spouses that are, on average, more educated than members of the average HH couple. More importantly for human capital accumulation, the mother in the HW marriages (those with children most likely to have a Hispanic-sounding last name—are more educated than their WH peers.) Since mothers could be more important in the child’s human capital accumulation and education, this would suggest that a child with a Hispanic last name should complete more years of education  \autocite{gould2020does}.

@@

I provide the results of the estimation of equation \ref{eq:1a} in Table \ref{tab:lastname-ed-reg}. I estimate the mean educational outcomes of White, US-born Hispanics ages 25-40. I also restrict the sample to children of HW and WH parents. The omitted group is children of WH parents. Column 1 in Table \ref{tab:lastname-ed-reg} is the difference in total years of education between HW children and their WH peers. Column 2 is the difference in the probability of not completing high school. Column 3 is the difference in the probability of having an associate degree. Column 4 is the difference in the probability of having a bachelor’s degree. All regressions include controls for age, parental education and income, and state-year fixed effects.

@@

Although there is a small but significant gap in total years of education between HW and WH children, there is no significant difference in the probability of dropping out of high school. The gap between HW and WH high school dropouts is statistically insignificant (1 percentage point). The same is true for HW women (2 percentage points) and men (0 percentage points).

Notable differences emerge for higher education outcomes. HW children are 2 percentage points less likely to earn an associate degree and 3 percentage points less likely to earn a bachelor’s degree compared to their WH peers. HW children are 2 percentage points less likely to earn an associate degree compared to their WH peers, representing a 13.3\% reduction relative to the WH associate degree rate of 15\%. HW children are 3 percentage points less likely to earn a bachelor’s degree compared to their WH peers, a 13\% reduction compared to the WH bachelor’s degree rate of 23\%. These differences are slightly larger for HW women, who are 3 percentage points less likely to earn an associate degree and 4 percentage points less likely to earn a bachelor’s degree. For HW men, the gap is 2 percentage points for an associate degree and statistically insignificant for a bachelor’s degree.

>>>>

A large body of literature provides evidence of substantial earnings gaps across race and ethnicity in the US \autocite{bayer2018divergent, charles2008prejudice}. Hispanic individuals constitute a large and growing portion of the US population. As this population grows, it is crucial to determine whether ethnic discrimination affects their employment prospects, wages, and career advancement. Understanding these labor market outcomes is critical, as they directly affect broader societal issues like assimilation and economic mobility, \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017}. These factors serve as key indicators of how successfully Hispanic immigrants can navigate society and ascend the socioeconomic ladder.

In this paper, I answer the following questions. Does having a Hispanic last name affect educational outcomes? Does having a Hispanic last name affect labor market outcomes? I aim to show that comparing Hispanic White individuals to non-Hispanic White individuals might create an artificially higher earnings gap since these groups differ in many observable characteristics.\footnote{In this study, I define Hispanic individuals specifically as US-born persons who self-identify as White and have at least one parent born in a Spanish-speaking country.} \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} My analysis focuses on US-born children with one foreign-born parent (i.e., inter-ethnic).\footnote{This study's focus on US-born children with foreign-born parents means the findings may not generalize to Hispanic children with US-born parents, who likely face different socioeconomic and cultural circumstances. The analysis does not account for heterogeneity in immigrant characteristics across different Spanish-speaking countries of origin, such as variations in educational attainment, socioeconomic status, and gender-specific migration patterns, which could influence both parental selection into migration and subsequent child outcomes.} Others have attempted to compare how native-born White Hispanics fare compared to non-Hispanic Whites and foreign-born Hispanics. \cites{antman2020ethnic} have compared the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and non-Hispanic Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health.\footnote{For additional studies examining ethnic identity and outcomes among Hispanic populations, see \textcite{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, which collectively document the relationships between ethnic identification, assimilation patterns, and various socioeconomic outcomes.} \textcite{davilaChangesRelativeEarnings2008} found gaps in labor market outcomes between Hispanics and non-Hispanic Whites, which they attribute largely to differences in education, experience, immigration status, and regional differences. This paper builds on those studies by emphasizing how cultural assimilation and generational status further shape educational and labor market disparities for Hispanics.

Understanding discrimination against Hispanics in labor markets has far-reaching implications that extend beyond individual earnings. Labor market discrimination can create persistent intergenerational disadvantages by limiting economic mobility, reducing access to quality healthcare and education, and constraining residential choices; \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,bowles2002inheritance, djajic2003assimilation}. Discrimination may foster occupational sorting and segregation, as shown by recent task-based models of racial wage gaps that identify race-specific barriers in “Contact” tasks and explain the stagnation of these gaps post-1980 \autocite{hurst2024task}. These barriers may trap Hispanic families in cycles of lower socioeconomic status, as reduced earnings and employment opportunities limit their ability to invest in their children's human capital or build wealth through homeownership and savings.\footnote{See \textcite{bowles2002inheritance} for an analysis of the intergenerational transmission of economic status.}

@@

This study builds and improves upon the traditional Oaxaca-Blinder-Kitagawa decomposition approach often used in labor market discrimination studies \autocite{kitagawa1955components, oaxaca1973male,blinder1973wage}. While these decomposition methods have provided valuable insights into earnings gaps between groups, including Hispanics and non-Hispanic Whites \autocite{davilaChangesRelativeEarnings2008}, they are limited in their ability to account for unobserved differences between the groups. My approach comparing children of inter-ethnic marriages, provides a more closely matched comparison group, allowing me to better isolate the effect of likely having a Hispanic last name on labor market outcomes.

The methodological improvement in this paper is particularly timely and relevant given the rapidly changing demographics of the United States. The US population is growing increasingly diverse, with significant implications for labor market dynamics and potential discrimination. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019.\footnote{The proportion of non-Whites and Hispanics is based on the author's calculations using the Current Population Survey (CPS).} Native-born White Hispanic men earn 21\% less than non-Hispanic White men, although there is evidence that a substantial portion of this gap is due to differences in education \autocite{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. These discrepancies may also stem from discrimination against Hispanics, which can lead to reduced job opportunities, lower wages, and hindered assimilation \autocite{chettyWhereLandOpportunity2014,bowles2002inheritance, djajic2003assimilation}.\footnote{This study focuses on US-born children of inter-ethnic unions, many of whom have at least one foreign-born parent. Assimilation concerns remain relevant because foreign-born parents may pass on cultural norms and language preferences, which can still shape these children's educational and labor market outcomes.}

 Using this novel approach, I find significant effects of Hispanic surnames on both educational and labor market outcomes. Individuals who likely have Hispanic last names face substantial disadvantages in educational attainment and earnings compared to their counterparts with likely non-Hispanic names. Specifically, individuals with presumably Hispanic last names complete 0.2 fewer years of education than those with non-Hispanic names, even when controlling for family background. Individuals with presumably Hispanic last names are also 1 percentage point more likely to be unemployed and earn 5 percentage points less than those with likely non-Hispanic names. These findings suggest that ethnic discrimination continues to play a role in shaping economic opportunities for Hispanics in the United States, highlighting the need for targeted policies to address these persistent disparities.

@@

The main identifying assumption of my empirical strategy depends on the assumption that people born to HW parents are similar to their WH peer in all observable---and unobservable---aspects and characteristics relevant to the labor market, so surname is the only difference between these groups. Children from US-born-Hispanic homes might benefit from more favorable family conditions than those from Hispanic-US-born homes, considering Whites statistically have higher socioeconomic status than Hispanics.  These factors introduce doubts regarding whether children from Hispanic-US-born families receive comparable familial support and influences, genetically or environmentally, as those from US-born-Hispanic families.

@@

I use three datasets: the Integrated Public Use Microdata Series (IPUMS) Current Population Survey (CPS), CPS's Annual Social and Economic (ASEC) supplement, and CPS's outgoing rotation \autocite{cps2019}, and the 1960 to 2000 US censuses \autocite{acs2019}.

I use the 1994–2019 CPS data set to study the effect of Hispanic surnames on labor market outcomes, because this time period contains data on parents’ place of birth, ethnicity, and race. The CPS does not provide data on parents’ characteristics, essential to determine the family background, but the 1960–2000 US Census data includes parents’ place of birth, race, and ethnicity. I employ this information to construct 'synthetic parents' using a method developed by \textcite{rubinstein2014pride} linking husbands and wives  in the census data. Assuming that parents have children 25–40 years old, I link these “parents” using parents’ places of birth, birth year of "children," and the parents’ places of birth.

@@

I use the CPS for my primary analysis of the effect of a Hispanic surname on earnings. I restrict my sample to US citizens  aged 25–40 born between 1960 and 2000 in the US. Taking advantage of data on parents’ place of birth, I divide the sample into four groups depending on their parents’ ethnicity. Mothers or fathers are Hispanic if they were born in a Spanish-speaking country or Puerto Rico, and non-Hispanic White if they were born in the United States.\footnote{The list of Spanish-speaking countries includes: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} Therefore, an observation can be the product of four types of parents:
\begin{enumerate}
\item US-born father and US-born mother (hereafter WW) 
\item US-born father and Hispanic mother (hereafter WH)---the comparison group and the group that is likely to have a non-Hispanic-sounding last name
\item Hispanic father and US-born mother (hereafter HW)---the focus of this study and the group most likely to have a Hispanic-sounding last name
\item Hispanic father and Hispanic mother (hereafter HH).
\end{enumerate}

@@

Regarding the classification of US-born parents as 'White'. While the United States has a long history of immigration, the probability that a US-born parent in this sample is a second-generation or later immigrant from a Spanish-speaking country is very low, as shown in Table \ref{tab:mat3}. The majority of Hispanics in the US during this period were first- and second-generation immigrants \autocite{antmanEthnicAttritionAssimilation2020}. Only 3\% of native-born Americans identified as Hispanic during this period, making it statistically unlikely that an interethnic child with a native-born parent is also a second-generation Hispanic immigrant. While this approach limits the generalizability of the findings to the broader Hispanic population, it provides a cleaner identification of surname effects, which is the primary focus of this study.

@@

Using the 1960 to 2000 censuses, I constructed a data set of synthetic parents. The sample includes married White men and women. Even though the census asks a person whether they are Hispanic or not, I took advantage of the questions on place of birth to create a proxy for ethnicity. I consider a Hispanic persons who self-report White as their racial group and are born in a Spanish-speaking country. Consequently, US born in the sample are those who self-report White as race and are native-born. Using the information provided in the census, I can link husbands and wives with each other. I assume that parents have children between the ages of 25 and 40, so my sample consists of married White men and women with children born in the 1920 to 1975 cohorts\footnote{The construction of ``synthetic parents'' follows the method used by \textcite{rubinstein2014pride}.}.

To illustrate the construction of synthetic parents more concretely, consider someone who was 35 years old in 1999, meaning they were born in 1964. If this person's mother was born in Mexico and their father was born in the United States, their 'synthetic parents' would be identified using the 1970 Census data, when the person was 6 years old. The 'synthetic mother' would have the average characteristics (education, income, etc.) of Mexican-born women who were married to US-born men and had children around 1964, when they were between 20 and 35 years old (meaning they were born between 1929 and 1944). Similarly, the 'synthetic father' would have the average characteristics of US-born men who were married to Mexican-born women and had children in that same year.

This aggregation process introduces some heterogeneity, as individual parents vary considerably from these group-level measures. However, this approach provides valuable information about typical family background characteristics that would otherwise be completely unobserved. The synthetic parents' characteristics serve as useful proxies for the socioeconomic environment in which these children were raised, allowing me to control for important background factors that influence educational and labor market outcomes. While more refined matching would be ideal, the publicly available data limit the potential matching dimensions. Nevertheless, the current approach provides meaningful estimates while acknowledging these data constraints.

@@

Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variable equal to 1 if person $i$ has a Hispanic father and US-born mother, and 0 if person $i$ has a US-born father and Hispanic mother. $X_{ist}$ is a vector of controls that includes age and hours worked, $\gamma_{st}$ represents state-year fixed effects, and $\phi_{ist}$ is the error term. The equation for this strategy is written as follows, and the sample is restricted to individuals from WH and HW families:

\begin{equation} \label{eq:1a}
Y_{ist} = \beta_{1} HW_{ist} + X_{ist} \pi + \gamma_{st} + \phi_{ist}
\end{equation}

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gap in outcomes between children of inter-ethnic marriages who likely have a Hispanic-sounding last name versus those who likely have a White-sounding last name. If $\beta_{1} > 0$, then people who likely have a Hispanic last name have better outcomes than people who likely have a White last name. If $\beta_{1} < 0$, then people who likely have a Hispanic last name have worse outcomes than people who likely have a White last name.

By comparing children from interethnic marriages, I can control for many unobservable family characteristics that may affect labor market outcomes, since these families are likely to be more similar to each other than to families with two Hispanic or two non-Hispanic White parents. This approach allows me to isolate the effect of surname, providing a more accurate estimate of its impact. The difference in means between Hispanic and non-Hispanic Whites could result from discrimination, but it could also be due to differences in innate abilities, skills, and parental investments . WH children will likely have a non-Hispanic last name, while HW children will likely have a Hispanic last name. This method was developed by \textcite{rubinstein2014pride}. This approach improves upon traditional Oaxaca-Blinder-Kitagawa decomposition methods, which are limited in their ability to account for unobserved differences between groups, by providing a more suitable comparison group to measure discrimination \autocite{oaxaca1973male,blinder1973wage,kitagawa1955components}.

@@

Two primary reasons underscore this assumption: 
\begin{itemize}
    \item First, there is significant selection in the marriage market. Since belonging to the Hispanic out-group may bring negative societal effects, it is worth exploring factors common among White women who have children with Hispanic men. 
    \item Second, fathers and mothers influence human capital accumulation differently \autocite{kimball2009risk,magruder2010intergenerational}. If marriage were random, then WH households might be a better environment for children. Using the CPS and the US Census, I will evaluate these empirical concerns.
\end{itemize}

A large body of empirical and theoretical work indicates that marriages exhibit strong selection, or `assortative matching' \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996, beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}. Also, \textcite{duncanIntermarriageIntergenerationalTransmission2011} show a similar pattern of assortative matching among Mexicans in the United States. Interethnic WH and HW marriages generally consist of partners sharing similar characteristics, suggesting both WH and HW parents are similar in aspects relevant to labor market outcomes.

Another threat to identifying a true effect could arise from measurement error in using parents’ place of birth from the CPS data as a proxy for a parent’s ethnicity and last name: The CPS notes place of birth but not ethnic or racial identity. It is possible that a native-born father could be a second-generation or later immigrant from a Spanish-speaking country, but this is unlikely. Most Hispanics from 1960 to 2000,  were first-generation immigrants, and the number of second-generation or later was very small (Table \ref{tab:mat3}).\footnote{Based on the author's calculations using US Census data.} Only 3\% of native-born Americans identified as Hispanic, making it unlikely that an interethnic child with a native-born father is also a second generation Hispanic immigrant.

It is important to address potential statistical inference issues related to using synthetic parents' characteristics as control variables. Regarding the 'generated regressor' issue: The parental characteristics are included solely as controls and not for inference purposes. I do not interpret these coefficients, as my primary interest lies in $\beta_1$, which captures the gap between Hispanic and White last names. Since I am only using these synthetic characteristics as controls and not making inference about their coefficients, this should not affect the standard errors or p-values of my coefficient of interest $\beta_1$, which is not a generated regressor.

The synthetic parents are constructed using the birth year of the child and are not based on actual parents in the CPS sample at the time of the survey. The place of birth of parents—mothers and fathers—are questions asked of all participants in the Current Population Survey (CPS) starting 1994. Therefore, I do not use the information of parents who are in the CPS sample at the time of the survey, but rather construct synthetic parents from historical Census data. This approach means that concerns about non-random attrition of parents from the CPS sample do not apply to my analysis.

Nevertheless, measurement error may still arise from the use of group averages that may not perfectly represent the actual parents of each individual in my sample. However, using group means typically reduces measurement error compared to individual-level measures. Since these variables are used as controls rather than variables of primary interest, any remaining measurement error is unlikely to substantially bias the coefficient of interest ($\beta_1$) on Hispanic last name, though it might reduce the precision of the estimates.

I cluster standard errors at the state level. This approach accounts for correlation in the error terms within geographical regions, which is likely to capture much of the potential correlation structure. This clustering approach is common in the literature and helps ensure that the statistical significance of the results is not overstated due to correlated errors.

@@

It is important to acknowledge potential threats to identification in this approach. While HW and WH couples exhibit similar observable socioeconomic characteristics, they may differ systematically in unobservable dimensions that influence labor market outcomes for their children. These unobservable differences could include parenting styles, educational preferences, time investments in children, gender norms, or cultural attitudes that are not captured in conventional socioeconomic metrics.

However, this approach offers distinct advantages over traditional Oaxaca-Blinder-Kitagawa decomposition methods. First, by focusing specifically on children from interethnic marriages with similar parental characteristics but different surname ethnicities, I use a more targeted comparison group that better isolates the effect of perceived ethnicity from family background. Traditional decomposition approaches would struggle to separate discrimination effects from the broader socioeconomic gaps between Hispanic and White families overall. Second, the smaller educational and income disparities between HW and WH families compared to HH and WW families suggest that selection bias is less severe in this comparison, allowing for a clearer attribution of outcome differences to discrimination rather than unobserved heterogeneity. Nevertheless, I acknowledge that residual selection bias remains a concern, particularly regarding cultural and attitudinal factors that may systematically differ between HW and WH households despite similar educational and economic profiles.

Using the “synthetic” parents I constructed using the Census data, I examine the family background of the different types of children. WW couples have the highest level of education among the four groups: 12.58 years  for husbands and 12.36 for wives, so a WW household has 24.95 years of schooling. Men in HH marriages have 8.91 years of education, while women have 8.68. As a household, HH couples have 17.69 years of education. As predicted, interethnic couples marry people similar to them. WH husbands have 11.82 years of education, while wives have an average of 10.71, a household total of 22.68 years. HW husbands have 10.33 years, while wives have an average of 11.01, giving HW households a total of 21.50 years. Both  Hispanic men and women in interethnic couples marry white spouses who are, on average, more educated than members of the average HH couple. More importantly for human capital accumulation, the mother in the HW marriages (those with children most likely to have a Hispanic-sounding last name—are more educated than their WH peers.) Since mothers could be more important in the child’s human capital accumulation and education, this would suggest that a child with a Hispanic last name should complete more years of education  \autocite{gould2020does}.

@@

I present the results from estimating equation \ref{eq:1a} in Table \ref{tab:lastname-ed-reg}. I estimate the mean educational outcomes of White Hispanic, US-born individuals ages 25-40. I also restrict the sample to children of HW and WH parents, and the omitted group is children of WH parents. Column 1 in Table \ref{tab:lastname-ed-reg} is the difference in total years of education between HW children and their WH peers. Column 2 is the difference in the probability of not completing high school. Column 3 is the difference in the probability of having an associate degree. Column 4 is the difference in the probability of having a bachelor’s degree. All regressions include controls for age, parental education and income, and state-year fixed effects.

@@

Although there is a modest yet statistically significant gap in total years of education between HW and WH children, there is no significant difference in the probability of dropping out of high school. The gap between HW and WH high school dropouts is statistically insignificant (1 percentage point). The same is true for HW women (2 percentage points) and men (0 percentage points).

Notable differences emerge for higher education outcomes. HW children are 2 percentage points less likely to earn an associate degree compared to their WH peers, representing a 13.3\% reduction relative to the WH associate degree rate of 15\%. Similarly, they are 3 percentage points less likely to earn a bachelor's degree, a 13\% reduction compared to the WH bachelor's degree rate of 23\%. These differences are slightly larger for HW women, who are 3 percentage points less likely to earn an associate degree and 4 percentage points less likely to earn a bachelor’s degree. For HW men, the gap is 2 percentage points for an associate degree and statistically insignificant for a bachelor’s degree.
